---
title: 'W203 Lab 1: Candidate Dept EDA'
author:
- Jon Mease
- James De La Torre
- Adam Yang
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

As members of the campaign committee for the upcoming Washinton state election, we are interested in the nature of debt that candidates for office have taken on in past elections. This report contains the findings of our initial exploratory data analysis of a dataset covering the 2012 election.

### Research question
How can the amount of debt taken on by candidates in the 2012 Washington state election be understood in relation to all of the other available fields in this dataset?

### Dataset

### Load your data set into R.
First we load the data set, treating the string `'#N/A'` as a missing value.
```{r}
DebtRaw = read.csv("../dataset/CandidateDebt.csv", na.strings='#N/A')
```
Initially is appears that we have 1043 observations over 28 variables.

Next we examine the first 4 rows of the dataset
```{r}
str(DebtRaw[1:4,])
```

From this information we can see that there appears to be a mis-alignment between the column headers and the datasets starting around the 10th column `legislativedistrict`.

Then insert a new DUMMY column header in between the 9th and 10th column, then display `DebtShifted`

```{r}
DebtShifted = DebtRaw
colnames(DebtShifted) <- c(colnames(DebtRaw)[1:9],
                           'DUMMY',
                           colnames(DebtRaw)[10:(ncol(DebtRaw)-1)])
str(DebtShifted[1:4,])
```

### Outline:
Outline of prose to cover the following prompts:

 - Describe your data set. What types of variables does it contain? How many observations are there?
 - Evaluate the data quality. Are there any issues with the data? Explain how you handled these potential issues.
 - Explain whether any data processing or preparation is required for your data set.

Discuss the type and meaning of the variables (including DUMMY), Note that DUMMY has nearly the same factors as office

Then examine some of the rows (example of one that lines up and one that does not)

Then example of one that appears to be out of order (Jill Johnson)

Then discussion of the repititions in the right. But amount always repeats with all of the other columns.

It seems that left and right are two tables that are not (generally) aligned. Given rows that align and Jill Johnson

Speculate that some form of malformed SQL join introduced the duplicates.

We will look at the right table only, with duplicates removed. Note that this assumes that we don't have valid identical transactions on identical days.

```{r}
# Split off left table
DebtLeft = DebtShifted[, seq(9)]

# Split off right table
DebtRight = DebtShifted[, seq(10, ncol(DebtRaw))]

# Rename DUMMY to office
colnames(DebtRight)[[1]] <- "office"

# Keep cases with at least one non-NA entry
DebtRightValid <- DebtRight[rowSums(!is.na(DebtRight)) > 0,]

# Drop duplicates
Debt <- unique(DebtRightValid)
```

After cleaning steps above, we're left with 194 observations over 19 variables


## Univariate Analysis of Key Variables

Univariate analysis...

## Analysis of Key Relationships

Key relationships...

### How much money is each party spending?

The first bivariate relationship explored was the total expenditure by political party.  The working debt dataset was grouped by party, and the sum of the dollar amounts for each group was computed.  Figure 10 shows a bar chart with this information.  Expenditures in the dataset were overwhelmingly from Democrats,  accounting for 79% of the total dolalr amount.

```{r}

# sum amount by party
amountbyparty <-aggregate(Debt['amount'], by=list(party=Debt$party), FUN=sum)

#format method, which is necessary for formating in a data.frame   
format.money  <- function(x, ...) {
  paste0("$", formatC(as.numeric(x), format="f", digits=2, big.mark=","))
}
class(amountbyparty$amount) <- c("money",class(amountbyparty$amount))


# sort the list descending, create pareto
amountbyparty <- amountbyparty[order(amountbyparty$amount, decreasing=TRUE),]

# bar plot tutorial https://www.statmethods.net/graphs/bar.html
barplot(amountbyparty$amount, main="Total Expenditures by Political Party", 
  	xlab="Political Party",
  	names.arg=amountbyparty$party,
  	cex.names=1)

```




### Where is the money going?

The next relationship explored was expenditure by expense category.  The dataset's description column provided some categorization of expenses, but since the description field was populated by different people for different reports, there was not much consistency of description values.  Out of 194 records, there were 105 unique values in the description column.  Another column in the dataset, code, was supposed to provide a more generalized categorization of expenses.  However, this field was populated in only 94 of 194 records and only encompassed 20% of the grand total of all expenses in the dataset ($55,326.09 out of $269,191.70).

Since the description data were too granular, and since the code data were too sparse, a new column, coarsedescription, was generated by manually examining the values in the original description column and assigning more general, coarse labels to them.  For example, there were 23 description values of the form "TREASURY " + month.  These were all assigned into coarsecategory "TREASURY".  Similarly, there were several descriptions values that indicated they were related to consulting, so these were assigned coarsedescription "CONSULTING".  In some cases where the nature of an expense was not immediately clear, a quick bit of web searching for the type of business in the vendorname column helped to determine a reasonable value for coarsedescription.  In those cases, an explanation of the reasoning for selecting the value of coarse category is provided in a comments column.  There is the possibility that some description values did not get mapped into their optimum coarsedescription category, but the author is reasonably confident that the mapping is sensible.  The 106 unique description values were mapped to 23 coarsedescription values.  For a complete listing of the translation from description to coarsedescription, see Appendix A.

A pareto of total expenditure by coarsedescription is shown in Figure 10.  The total for a given coarsedescription is computed simply by summing the values in the amount column for all records having that coarsedescription.

```{r}
desc_to_coarsedesc = read.csv("../dataset/description_to_coarsedescription.csv", na.strings='#N/A')
# we need to merge coarsedescription into Debt by description
Debt2=merge(Debt,desc_to_coarsedesc[c('description','coarsedescription')])

# Put 0 for position when there is no value
Debt2$position[is.na(Debt2$position)] <- 0
Debt2$legislativedistrict[is.na(Debt2$legislativedistrict)] <- 0

# sum amount by coarsedescription
amountbycoarsedesc <-aggregate(Debt2['amount'], by=list(coarsedescription=Debt2$coarsedescription), FUN=sum)

#format method, which is necessary for formating in a data.frame   
format.money  <- function(x, ...) {
  paste0("$", formatC(as.numeric(x), format="f", digits=2, big.mark=","))
}
class(amountbycoarsedesc$amount) <- c("money",class(amountbycoarsedesc$amount))


# sort the list descending, create pareto
amountbycoarsedesc <- amountbycoarsedesc[order(amountbycoarsedesc$amount, decreasing=TRUE),]
#print(amountbycoarsedesc)

nbars=14
topn=amountbycoarsedesc[1:nbars,]
?barplot
# bar plot tutorial https://www.statmethods.net/graphs/bar.html
barplot(topn$amount, main="Total Expenditures by Coarse Description",
        sub=paste(c("(Top", nbars, "shown)"), collapse = " "),
  	xlab="Coarse Description",
  	names.arg=topn$coarsedescription,
  	cex.names=0.65)

# Maybe I should make a function for bar charts, since I will be making many of them.

```

### Do the parties spend their money differently?

The expenditures by category can be broken down by political party to see differences in how Republicans and Democrats allocate their expenses.  Since the total expenditures were so different between the parties, rather than look at the raw dollar amounts, the data were normalized to fraction of total expenditure for a given party.  Since there were only $102.88 in expenditures for Independent candidates (0.04% of total), all spent in a single coarse category (Mailing Services), Indpendent party data were excluded from this specific analysis.

Figure 12 shows a pareto of fraction of party total expenditure by coarse description, by party.  The data reveal that Democrats and Non-partisan candidates spent a greater fraction of money on consulting than the Republicans, while Republicans spent a greater fraction on credit card payments and telecom expenses.  Unfortunately, there is no visibility into what purchases were made using the credit cards (the raw description column just named the credit card (e.g., AMEX), so it provides no additional information, either.)

```{r}

# Rename "subject" column to "N"
names(amountbyparty)[names(amountbyparty)=="amount"] <- "totalamountbyparty"

#rollup debt2 by coarsedesc and party
expense_by_coarsedesc_by_party <-aggregate(Debt2['amount'],                                         by=list(coarsedescription=Debt2$coarsedescription, party=Debt2$party), FUN=sum)


#merge totalamount into expense_by_coarsedesc_by_party
frac_by_coarsedesc_by_party <- merge(expense_by_coarsedesc_by_party,amountbyparty)
# compute fraction
frac_by_coarsedesc_by_party$fractionofpartytotal <- frac_by_coarsedesc_by_party$amount / frac_by_coarsedesc_by_party$totalamountbyparty

# drop independents
frac_by_coarsedesc_by_party <- frac_by_coarsedesc_by_party[frac_by_coarsedesc_by_party$party != "INDEPENDENT", ]

# get max frac by coarsedescription (across parties)
maxfrac_by_coarsedesc <-aggregate(frac_by_coarsedesc_by_party['fractionofpartytotal'],
                                           by=list(coarsedescription=frac_by_coarsedesc_by_party$coarsedescription), FUN=max)

# Rename "fractionofpartytotal" column to "maxfracofpartytotal"
names(maxfrac_by_coarsedesc)[names(maxfrac_by_coarsedesc)=="fractionofpartytotal"] <- "maxfracofpartytotal"

#merge maxfrac into frac_by_coarsedesc_by_party
frac_by_coarsedesc_by_party2=merge(frac_by_coarsedesc_by_party,maxfrac_by_coarsedesc)


# assign part number to put nonpartisan last
frac_by_coarsedesc_by_party2$partynumber<-ifelse(frac_by_coarsedesc_by_party2$party=="DEMOCRAT",1,
		ifelse(frac_by_coarsedesc_by_party2$party=="REPUBLICAN",2,3
		))

#Now, sort the df by maxfracofpartytotal desc, then by partynumber
frac_by_coarsedesc_by_party2 <- frac_by_coarsedesc_by_party2[order(-frac_by_coarsedesc_by_party2$maxfracofpartytotal,
                                                                   frac_by_coarsedesc_by_party2$partynumber),]

# Now we can finally make a grouped pareto!
frac_by_coarsedesc_by_party2$barlabels <- ""
frac_by_coarsedesc_by_party2$barlabels <- frac_by_coarsedesc_by_party2$barlabels <-

  
# Trick Make labels which are non blank only when row mod 3 =2
x<-1:nrow(frac_by_coarsedesc_by_party2)
y <- x%%3==2

cd=frac_by_coarsedesc_by_party2$coarsedescription
frac_by_coarsedesc_by_party2$barlabels <- with(frac_by_coarsedesc_by_party2, ifelse(y,cd,"" ) )

cd <- frac_by_coarsedesc_by_party2$coarsedescription
frac_by_coarsedesc_by_party2$barlabels <- ifelse(y,frac_by_coarsedesc_by_party2$coarsedescription,"" )


# now make a sidebyside bar chart
barplot(frac_by_coarsedesc_by_party2$fractionofpartytotal, main="Fractional Expenditures by Coarse Description, by Party",
        xlab="Coarse Description", col=c("blue","red","green"),
        legend = c("Democrat","Republican","Non Partisan"), beside=TRUE, names.arg=frac_by_coarsedesc_by_party2$coarsedescription, las=2)

```

### For Which Types of Offices is the Most Money Spent?

An analysis of amount of expenditure by office was performed by summing the amounts spent by unique value of the office2 column (the office column from the right half of the dataset).  However, since the number of offices of a given type varied (there are more state representatives than state senators or governors), and since the number of candidates varied by office type, the amounts were divided by the number of unique combinations of legislativedistrict, position, and party within each office2 value.  The intent of this was to normalize the amounts to a "per campaign" value.  For example, there were 39 unique candidates for state representative.  In Washington, there are two state representatives per legislative district, and the position column designates which of these two slots a row belongs to.  The party column provides another level of granularity for a given line item.  A pareto of expenditures per campaign by office is shown in Figure 13.  Note that since the dataset does not have a unique candidate identifier, such as a number or registration ID, multiple candidates of the same party with expenditures for the same office and position would be treated as a single candidate.

The pareto shows that the race for governor had the greatest expenditure per candidate.  Two other statewide offices, treasurer and attorney general, showed the next highest expenditures per candidate.

```{r}
# get sum of amount by DUMMY
#sumbyoffice <- aggregate(Debt2['amount'],
#by=list(office=Debt2$DUMMY), FUN=sum)

# get sum(amount) and n rows by unique value of DUMMY, legislativedistrict, position, and party,
# and jurisdictiontype (in case I summarize by that later).
sumbycand <- aggregate(Debt2['amount'],
by=list(office=Debt2$office, legislativedistrict=Debt2$legislativedistrict, position=Debt2$position, party=Debt2$party, jurisdictiontype=Debt2$jurisdictiontype), FUN=sum)


# now sum the amounts in sumbycand and count rows by office
sumbyoffice <- aggregate(sumbycand['amount'], by=list(office=sumbycand$office), FUN=sum)
countbyoffice <- aggregate(sumbycand['position'], by=list(office=sumbycand$office), FUN=length)

# Rename cols
names(sumbyoffice)[names(sumbyoffice)=="amount"] <- "totalamount"
names(countbyoffice)[names(countbyoffice)=="position"] <- "ncandidates"

# merge the sum and count data
amountperjob <- merge(sumbyoffice,countbyoffice)
amountperjob$mean_amount <- amountperjob$totalamount / amountperjob$ncandidates

# sort by mean_amount desc
# sort the list descending, create pareto
amountperjob <- amountperjob[order(amountperjob$mean_amount, decreasing=TRUE),]

# bar plot tutorial https://www.statmethods.net/graphs/bar.html
barplot(amountperjob$mean_amount, main="Expenditure per Candidate by Office", 
  	xlab="Office",
  	names.arg=amountperjob$office,
  	cex.names=1, las=2)


```


### Do Different Jurisdiction Types Have Different Transaction Sizes?

Campaigns for offices of different jurisdiction types are likely conducted differently.  The dataset contains four unique values in its jurisdictiontypes column--judicial, legislative, local, and statewide.  Box plots of transaction amounts by jurisdictiontypes are shown in Figure 13.  To provide insight into the relative number of data points comprising each jurisdition type, the width of the boxes are drawn proportionally to the square root of the number of points.  The plot shows that while there were several outlier points for legislative offices, the highest amounts were in the outliers in the statewide offices.  The three outlier expenditure amounts for statewide offices were all for consulting expenses, whereas the outliers for legislative offices spanned many categories, including mailing services, graphic design and printing, telecom, consulting, and web and data services.  The single outlier for judicial offices was also consulting.  The single outlier expense for local offices was for a credit card payment (to Cabela's, a hunting, fishing, and caping supply store, oddly enough).

The small dataset size of 194 total points, and the disparate number of points by jurisdiction type (Judicial: 14; Legislative: 125; Local: 19; and Statewide: 36) may limit the insight obtained by the box plots, but examination of the outlier points give some indication that consulting fees are some of the most expensive transactions in political campaigns for most types of political offices in Washington.



```{r}

?boxplot
# Expenditure amounts by jurisdictiontype
boxplot(amount~jurisdictiontype,data=Debt2, main="Transaction Dollar Amounts by Jurisdiction Type", 
  	xlab="Jurisdiction Type", ylab="Amount in Dollars", varwidth=T)

```

### Do Expenditures Change with Time?

As a final investigation, the change in spending amounts over time was examined.  The thrudate value was chosen as the time value to use, since that is when expenditures are reported and since it groups expenditures by month.  Trend plots of the sum, count, and average expenditure amount vs. thrudate are shown in Figures 15-17, respectively.  The trends show that both the number of expenditures and the dollar amount of individual expenditures generally increased as the election drew near.

```{r}
# Time component

```


#### Just what are "Treasury" Expenditures?

Figures 15-17 show expenditures attributable to the 2012 election as far back as 2009.  These may or may not be erroneous data.  Investigation of the descriptions of expenditures in the earliest months indicate that the majority of expenditures through February of 2011 were described as Treasury expenditures.  To understand the significance of these expenditures, a dummy variable, Treasury, was created which was set to 1 for treasury expenses and 0 for all other categories.  The mean of this dummy variable by thrudate (the average fraction of individual expenditures in a month attributable to Treasury) was plotted vs. thrudate.  The plot shows that after February of 2011, the fraction of expenditures with this description decreases significantly, as other expense categories become more numerous.  There are 57 records of Treasury expenses in the dataset, of which 54 have the value "Operations and Overhead" in the code column.  The average amount of treasury expenses is $278.  Without further domain knowledge, one cannot know the exact meaning of these expenses, but perhaps these are monthly fees paid for the services of each campaign's treasurer.


```{r}
# Time component

# Copy Debt2 to Debt3
Debt3 <- Debt2

# Create dummy variable, TREASURY
Debt3$Treasury<-ifelse(Debt3$coarsedescription=="TREASURY",1,0)


" HERE I AM END OF EVENING 24MAY2018.  I would like to 
know how to perform multiple aggregations on multiple variables in R.

I have thrudate, amount, and Treasury.  I need sum, count, and mean of amount and mean of Treasury by thrudate

"

# Now get sum, count, and mean of amount and mean of Treasury by thrudate
#timestats=aggregate(Debt3$amount, function(x) {
# c(sum=sum(x), count=length(x), mean=mean(x))}, by=list(thrudate=Debt3$thrudate))

#timestats <- aggregate(. ~thrudate, data = Debt3[c("thrudate", "amount")], FUN = function(x) c(sm=sum(x), ave = #mean(x), n = length(x) ) )



# sum amount by party
#amountbyparty <-aggregate(Debt['amount'], by=list(party=Debt$party), FUN=sum)




#merge(aggregate(z~y, data[data$x==1,], function(x) {
# c(mean=mean(x, na.rm=TRUE), quantile(x, na.rm=TRUE))
#}), list(y=unique(data$y)), all=TRUE)

#plot(duration, waiting,            # plot the variables 
#+   xlab="Eruption duration",        # x−axis label 
#+   ylab="Time waited")              # y−axis label

```


### What jurisdiction types to parties spend their money on?

Interesting observations:
 - Democrats outspent all others on Legislative and Statewide elections
 - Republicans focused on Legislative and Local elections
 - Non partisians outspent all others on Judicial and Local elections

```{r}
party_jurtype_sum <- aggregate(Debt['amount'],
                                  by=list(party=Debt$party,
                                          jurisdictiontype=Debt$jurisdictiontype),
                                  FUN=sum)

# Order by total spending
xtab_sum <- xtabs(amount ~ jurisdictiontype + party, data=party_jurtype_sum)
xtab_sum <- xtab_sum[,order(colSums(xtab_sum), decreasing = TRUE)]

xtab_sum_mat <- data.matrix(xtab_sum)
barplot(xtab_sum_mat,
        beside = TRUE,
        legend.text = rownames(xtab_sum_mat),
        main = "Total party debt across jurisdiction types",
        xlab = "party",
        ylab = "total debt amount")
```


## Analysis of Secondary Effects ...

Secondary effects...

Jurisdiction Types vs Party counts?

Party vs Office counts?

## Conclusion

Debt amount is highly skewed in the positive direction.

Statewide amounts have the largest skew of the jurisdiction types

Total expendatures for democrats far exceed those for all other parties combined


Conclusion...