---
title: "W203 Lab 1: Candidate Dept EDA"
output: html_document
author: 
 - Jon Mease
 - James De La Torr
 - Adam Yang
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is the introduction...

### State the research question that motivates your analysis.

### Load your data set into R.

```{r}
DebtRaw = read.csv("../dataset/CandidateDebtDummy.csv", na.strings='#N/A')

# Split off left table
DebtLeft = DebtRaw[, seq(9)]

# Split off right table
DebtRight = DebtRaw[, seq(10, ncol(DebtRaw))]

# Rename DUMMY to office
colnames(DebtRight)[0] <- 'office'

# Keep cases with at least one non-NA entry
DebtRightValid <- DebtRight[rowSums(!is.na(DebtRight)) > 0,]

# Drop duplicates
Debt <- unique(DebtRightValid)
```
### Describe your data set. What types of variables does it contain? How many observations are there?

### Evaluate the data quality. Are there any issues with the data? Explain how you handled these potential issues.

### Explain whether any data processing or preparation is required for your data set.


## Univariate Analysis of Key Variables

Univariate analysis...

## Analysis of Key Relationships

Key relationships...

## Analysis of Secondary Effects ...

Secondary effects...

## Conclusion
Conclusion...